A memory frontier for complex synapses: Supplementary material

### 2013-05-30 NIPS/SynapseNIPSsupp.tex
added supp

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
% arXiv Paper ************************************************
%
% Subhaneil Lahiri's template
%
% Before submitting:
%    Comment out hyperref
%    Comment out showkeys
%    Replace \input{mydefs.tex} with its contents
%       or include mydefs.tex in zip/tar file
%    Replace \input{newsymb.tex} with its contents
%       or include newsymb.tex in zip/tar file
%    Put this file, the .bbl file, any picture or
%       other additional files and natbib.sty
%       file in a zip/tar file
%
% **** -----------------------------------------------------------
\documentclass[12pt]{article}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
%
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%equilibrium distribution
\newcommand{\eq}{\mathbf{p}^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
\DeclareMathOperator{\SNR}{SNR}
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
% ----------------------------------------------------------------
\input{sl_theorems_preamble.tex}
% ----------------------------------------------------------------
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{Area maximisation in continuous time}
%
% Author List:
%
\author{Subhaneil Lahiri
%
}

@@

\begin{abstract}
  We try to find the continuous time Markov process that has the maximal area under the signal-to-noise curve.
\end{abstract}

\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Beginning of Article:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this section we'll provide a summary of all the relevant properties of ergodic Markov chains in continuous time.
It is a straightforward generalisation of material that can be found in \cite{kemeny1960finite} with some ideas from \cite{hunter2000survey}.

@@

This allows us to write the lumpability condition \eqref{eq:lump}, and the transition matrix for the lumped process compactly:
%
\begin{equation}\label{eq:lumpcompact}
  VU\MM V = \MM V,
  \qquad
  \widehat{\MM} = U\MM V.
\end{equation}
%
By induction, one can show that similar relations hold for all powers:
%
\begin{equation}\label{eq:lumppower}
  VU\MM^nV = \MM^nV,
  \qquad
  \widehat{\MM}^n = U\MM^n V,
\end{equation}
%
and, via the Taylor series, for the exponential as well:
%
\begin{equation}\label{eq:lumpexp}
  VU \e^{t\MM} V = \e^{t\MM}V,
  \qquad
  \e^{t\widehat{\MM}} = U\e^{t\MM} V.
\end{equation}
%
The equilibrium distribution of the lumped process is given by
%
\begin{equation}\label{eq:lumpeq}
  \widehat{\mathbf{p}}^\infty = \eq V.
\end{equation}
%

\section{Signal-to-Noise ratio (SNR)}\label{sec:SNR}

@@

The individual potentiation/depression events will be described by \emph{discrete}-time Markov chains:
%
\begin{equation}\label{eq:MWdef}
  \M^\pm \equiv \I + \W^\pm,
  \qquad
  \M^\pm\onev = \onev,
  \qquad
  \M^\pm_{ij} \in [0,1].
\end{equation}
%
The initial signal creation event occurs at time $t=0$, but all subsequent potentiation/depression events occur at random times according to Poisson processes with rates $rf^\pm$, where $f^++f^-=1$ are the fraction of plasticity events that are potentiating/depressing respectively.
This means that the ``forgetting'' process will be described by the \emph{continuous}-time Markov chain:
%
\begin{equation}\label{eq:forgetting}
  \MM = r \frg \equiv r\prn{f^+\W^+ + f^-\W^-}.
\end{equation}
%
We only require that this Markov chain is ergodic. The Markov chains described by $\M^\pm$ need not be.

We assume that the probability distribution starts in the equilibrium distribution \eqref{eq:equilibrium}.
During the initial signal creation, a fraction $f^+$ will change to $\eq\M^+$ and a fraction $f^-$ will change to $\eq\M^-$.
After this, probabilities will evolve according to \eqref{eq:rowsum}.

@@

As we ignore correlations between different synapses, the only quantity of interest will be the number of synapses with each strength.
The signal can be defined as the increase in synaptic strength due to potentiation plus the decrease in synaptic strength due to depression.
If there are $N$ synapses and the strength of the synapse in each state is given by the column vector, $\w$, this is given by
%
\begin{equation}\label{eq:signal}
  S(t) = N f^+ [\mathbf{p}(t|+)-\eq]\w - N f^+ [\mathbf{p}(t|-)-\eq]\w,
\end{equation}
%
where $\mathbf{p}(t|\pm)$ is the probability of being in each state conditional on potentiation/depression at $t=0$:
%
\begin{equation}\label{eq:initprob}
  \mathbf{p}(t|\pm) = \eq\M^\pm \e^{rt\frg}.
\end{equation}
%
We can rewrite the signal as
%
\begin{equation}\label{eq:signalr}
\begin{aligned}
  S(t) &= N \eq (f^+\W^+ - f^-\W^-)\e^{rt\frg}\w\\
    & N (2f^+f^-) \eq (\W^+ - \W^-)\e^{rt\frg}\w.
\end{aligned}
\end{equation}
%
For the noise, we can use the standard deviation of the synaptic strength:
%
\begin{equation}\label{eq:noise}
  \text{Noise} = \sqrt{N \prn{\sum_i \eq_i\w_i^2 - (\eq\w)^2}}.
\end{equation}
%
From now on we shall assume that there are only two possible synaptic strengths, strong and weak, which we will describe by $\w_i=\pm1$. This splits the set of states into two subsets:
%
\begin{equation}\label{eq:plusminusstates}
  \boldsymbol{+} = \set{k}{\w_k=+1},
  \qquad
  \boldsymbol{-} = \set{k}{\w_k=-1}.
\end{equation}
%
This simplifies the noise:
%
\begin{equation}\label{eq:noisepm}
  \text{Noise} = \sqrt{N \prn{4\eq_+\eq_-}}.
\end{equation}
%
The factors of $\eq_+\eq_-$ will have little effect on anything that follows, so we will drop them.\footnote{They will not affect the optimal transition rates for most quantities, and the optima will correspond to the symmetric case when $\eq_+=\eq_-=\half$.} This leaves the signal-to-noise ratio as
%
\begin{equation}\label{eq:SNRcurve}
  \SNR(t) = \sqrt{N} (2f^+f^-) \eq (\W^+ - \W^-)\e^{rt\frg}\w.
\end{equation}
%

We can express this in terms of the one parameter family of transition matrices:
%
\begin{equation}\label{eq:Walpha}
  \begin{aligned}
  \W(\alpha) &= \alpha\W^++(1-\alpha)\W^-,
  &\quad&\implies&
    \frg&=\W(f^+),\\&&&&
    \W^+ - \W^- &= \diff{\W}{\alpha},\\&&&&
    \eq \diff{\W}{\alpha} &= -\diff{\eq}{\alpha} \frg.
  \end{aligned}
\end{equation}
%
Then \eqref{eq:SNRcurve} becomes
%
\begin{equation}\label{eq:SNRalpha}
  \SNR(t) = \sqrt{N} (2f^+f^-) \diff{\eq}{\alpha}(-\frg)\,\e^{rt\frg}\w.
\end{equation}
%

\subsection{Degrees of freedom and constraints}\label{sec:constraints}

We would like to use the components of $\W^\pm$ as the independent degrees of freedom when we try to maximise quantities associated with the SNR curve \eqref{eq:SNRcurve}.
However, \eqref{eq:MWdef} imposes some restrictions on the space of allowed values.
We can solve some of these by setting
%
\begin{equation}\label{eq:Wdiag}
  \W^\pm_{ii} = -\sum_{j\neq i} \W^\pm_{ij}.
\end{equation}
%
Then we treat the off-diagonal components as the independent degrees of freedom.
These must satisfy the following inequalities:
%
\begin{align}\label{eq:Wmin}
  \W^\pm_{ij} &\geq 0 \qquad \text{for $i \neq j$,}\\
\label{eq:Wmax}
  \sum_{j \neq i} \W^\pm_{ij} &\leq 1.
\end{align}
%
The inequalities $\W^\pm_{ij} \leq 1$, $-1\leq\W^\pm_{ii}\leq0$ follow automatically from these.

In \sref{sec:rescale}, we will see that the area under the SNR curve is invariant under the scaling $\W^\pm\to\lambda\W^\pm$.
This means that we can ignore the constraint \eqref{eq:Wmax} when maximising the area:
after the area has been maximised, we can use this degree of freedom to enforce \eqref{eq:Wmax} without changing the area.
However, this scaling does change the SNR curve, in particular its initial value, so we can't ignore \eqref{eq:Wmax} when maximising that.

@@

Suppose that we have a partition such that $\W^+$ and $\W^-$ are simultaneously lumpable, and that all the states in each subset have the same synaptic strength (see \sref{sec:lump}):
%
\begin{equation}\label{eq:lumpablesynapse}
  VU\W^\pm V = \W^\pm V,
  \qquad
  VU\w=\w.
\end{equation}
%
We can define a new synapse with
%
\begin{equation}\label{eq:lumpedsynapse}
  \widehat{\W}^\pm = U\W^\pm V,
  \qquad
  \widehat{\w} = U \w,
  \qquad
  \widehat{\mathbf{p}}^\infty = \eq V.
\end{equation}
%
This synapse has an SNR curve:
%
\begin{equation}\label{eq:lumpedSNR}
  \begin{aligned}
    \frac{\SNR(t)}{\sqrt{N} (2f^+f^-)} &=  \widehat{\mathbf{p}}^\infty (\widehat{\W}^+ - \widehat{\W}^-)\e^{rt\widehat{\W}^F}\widehat{\w}. \\
      &= \eq VU (\W^+ - \W^-) VU \e^{rt\frg} VU \w. \\
      &= \eq (\W^+ - \W^-) VU \e^{rt\frg} VU \w. \\
      &= \eq (\W^+ - \W^-)\e^{rt\frg} VU \w. \\
      &= \eq (\W^+ - \W^-)\e^{rt\frg}\w. \\
  \end{aligned}
\end{equation}
%
\ie the lumped process has exactly the same SNR as the original one.

@@

Using the first line of \eqref{eq:signalr}, we can write the initial SNR as
%
\begin{equation}\label{eq:init}
  \frac{\SNR(0)}{\sqrt{N}} = I = ({\eq}^++{\eq}^-)(f^+\W^+-f^-\W^-)(\onev^+-\onev^-).
\end{equation}
%
Using $\W^\pm(\onev^++\onev^-)=0$ and \eqref{eq:compflux}:
%
\begin{equation*}
  r{\eq}^-(f^+\W^++f^-\W^-)\onev^+ = \F_{-+} = \F_{+-} = r{\eq}^+(f^+\W^++f^-\W^-)\onev^-,
\end{equation*}
%
we can rewrite \eqref{eq:init} as
%
\begin{equation}\label{eq:initflux}
  I = \frac{4\F_{-+}}{r} - 4{\eq}^+f^+\W^+\onev^- - 4{\eq}^-f^-\W^-\onev^+.
\end{equation}
%
The last two terms are guaranteed to be negative, as the diagonal parts of $\W^\pm$ cannot contribute.
Therefore
%
\begin{equation}\label{eq:initfluxineq}
  \SNR(0) \leq \frac{4\sqrt{N}\F_{-+}}{r}.
\end{equation}
%
This inequality is saturated if potentiation never takes it from a $+$ state to a $-$ state and depression never takes it from a $-$ state to a $+$ state.

\subsection{Maximising initial SNR}\label{sec:initmax}

The flux will be maximised if potentiation is guaranteed to take it to a $+$ state and depression is guaranteed to take it to a $-$ state:
%
\begin{equation}\label{eq:guarantee}
  \sum_{j\in+} \W^+_{ij} = 1 \quad \forall\, i\in-,
  \qquad
  \sum_{j\in-} \W^-_{ij} = 1 \quad \forall\, i\in+.
\end{equation}
%
This implies that the process is lumpable to a two state system (see \sref{sec:lump} and \sref{sec:SNRlump}):
%
\begin{multline}\label{eq:binarylump}
%\begin{aligned}
  \W^+ = \begin{pmatrix}
           -1 & 1 \\
           0 & 0 \\
         \end{pmatrix}
  ,\quad
  \W^- = \begin{pmatrix}
           0 & 0 \\
           1 & -1 \\
         \end{pmatrix}
  ,\quad
  \w = \begin{pmatrix}
           -1 \\
           1 \\
         \end{pmatrix}
  ,\\ \implies\quad
  \frg = \begin{pmatrix}
           -f^+ & f^+ \\
           f^- & -f^- \\
         \end{pmatrix}
  ,\quad
  \eq = \begin{pmatrix}
           f^- & f^+ \\
         \end{pmatrix}
  ,%&
%  \e^{rt\frg} &= \I+(1-\e^{-rt})\frg
%         \begin{pmatrix}
%           f^-+f^+\e^{-rt} & f^+(1-\e^{-rt}) \\
%           f^-(1-\e^{-rt}) & f^++f^-\e^{-rt} \\
%         \end{pmatrix}
%  ,
%  \end{aligned}
\end{multline}
%
which leads to the initial SNR %curve
\footnote{Note that including the dropped factors in \eqref{eq:noisepm} would only change this to
%$\SNR(t) = \sqrt{N (4f^+f^-)} \,\e^{-rt}$,
$\SNR(0) \leq \sqrt{N (4f^+f^-)}$,
which would have no effect on what follows.}
%
\begin{equation}\label{eq:binarySNR}
 % \SNR(t) = \sqrt{N} (4f^+f^-) \,\e^{-rt}.
  \SNR(0) \leq \sqrt{N} (4f^+f^-) .
\end{equation}
%
%The initial SNR is maximised at $f^+=f^-=\half$:
This is maximised at $f^+=f^-=\half$:
%
\begin{equation}\label{eq:initmax}
  \SNR(0) \leq \sqrt{N}.
\end{equation}
%

@@

The signal-to-noise curve is given by \eqref{eq:SNRalpha}.
%%
%\begin{equation}\label{eq:SNR}
%  \SNR(t) = \sqrt{N}(2f^+f^-) \diff{\eq}{\alpha} (-\frg) \e^{rt\frg} \w.
%\end{equation}
%%
The area is computed by integrating this
%
\begin{equation}\label{eq:area}
\begin{aligned}
  A &= \frac{\sqrt{N}(2f^+f^-)}{r} \diff{\eq}{\alpha} \brk{-\e^{rt\frg}}_0^\infty \w \\
    &= \frac{\sqrt{N}(2f^+f^-)}{r} \diff{\eq}{\alpha} (\I-\onev\eq) \w \\
    &= \frac{\sqrt{N}(2f^+f^-)}{r} \diff{\eq}{\alpha} \w.
\end{aligned}
\end{equation}
%

We can rewrite this using \eqref{eq:diffpT}, with $A=\sqrt{N}(2f^+f^-)\hat{A}$ and $\enc_{ij} \equiv \diff{\frg_{ij}}{\alpha}=\W^+_{ij}-\W^-_{ij}$
%
\begin{equation}\label{eq:areaT}
  \hat{A} = \sum_{i,j,k} \eq_i \enc_{ij} (\fptb_{ik} - \fptb_{jk}) \eq_k \w_k.
\end{equation}
%

@@

\subsection{Derivatives \wrt \texorpdfstring{$\W^\pm$}{W+/-}}\label{sec:deriv}

As discussed in \sref{sec:constraints},
we will regard the off-diagonal elements of $\W^\pm_{ij}$ to be the independent variables,
with $\W^\pm_{ii}=-\sum_{j \neq i} \W^\pm_{ij}$ imposed by hand.
Thus,
%
\begin{equation}\label{eq:basicderivs}
  \pdiff{\frg_{ij}}{\W^\pm_{gh}} = f^\pm \delta_{gi}(\delta_{hj}-\delta_{ij}),
  \qquad
  \pdiff{\enc_{ij}}{\W^\pm_{gh}} = \pm\delta_{gi}(\delta_{hj}-\delta_{ij}).
\end{equation}
%
The implicit $g \neq h$ that comes with all derivatives is unnecessary, as the derivatives above vanish when $g=h$.

In particular, differentiating \eqref{eq:funddef},
%
\begin{equation}\label{eq:derivZ}
  \pdiff{\fund_{ij}}{\W^\pm_{gh}} = rf^\pm \fund_{ig} (\fund_{hj}-\fund_{gj}).
\end{equation}
%
We can then differentiate expression \eqref{eq:areaZ} to get
%
\begin{equation}\label{eq:derivA}
  \begin{aligned}
    \pdiff{\hat{A}}{\W^\pm_{gh}} =&\,
      rf^\pm \sum_{ijkl} \enc_{ij} \w_k \pib_l \brk{
         \fund_{lg}(\fund_{hi}-\fund_{gi})(\fund_{jk}-\fund_{ik}) + \fund_{li}(\fund_{jg}-\fund_{ig})(\fund_{hk}-\fund_{gk}) } \\
    &\pm \sum_{kl} \w_k \pib_l \fund_{lg} (\fund_{hk} - \fund_{gk}) \\
    =&\,
      2rf^\pm \sum_{ij} \enc_{ij} \eq_i \eq_g \brk{
         (\fptb_{gi}-\fptb_{hi})(\eta^+_{i}-\eta^+_{j}) +
         (\fptb_{ig}-\fptb_{jg})(\eta^+_{g}-\eta^+_{h}) } \\
      &\pm 2 \eq_g (\eta^+_g - \eta^+_h) \\
    =&\,
      2rf^\pm \sum_{ij} \enc_{ij} \eq_i \eq_g
         (\fptb_{gi}-\fptb_{hi})(\eta^+_{i}-\eta^+_{j})
      \pm 2 \eq_g (\eta^+_g - \eta^+_h)\brk{1+\pdiff{\ln(\eq_g)}{\,\ln(f^\pm)}} \\
    =&\,
      2rf^\pm \eq_g \brk{\sum_i a_i (\fptb_{gi}-\fptb_{hi}) + c_g (\eta^+_{g}-\eta^+_{h})}
      \pm 2 \eq_g (\eta^+_g - \eta^+_h). \\
  \end{aligned}
\end{equation}
%
where $a_i$ and $c_k$ were defined in \eqref{eq:areacoeffs}.

>>>>

% -*- TeX -*- -*- US -*-
\documentclass{article} % For LaTeX2e
\usepackage{nips13submit_e,times}
\input{sl_preamble.tex}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figs/"}}
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%
%additional symbols:
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}
% ----------------------------------------------------------------
\input{sl_theorems_preamble.tex}
% ----------------------------------------------------------------

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{A memory frontier for complex synapses: Supplementary material}
%
% Author List:
%
\author{Subhaneil Lahiri and Surya Ganguli\\
Applied Physics Department, Stanford University, Stanford CA\\
\emaillink{sulahiri@stanford.edu}, \emaillink{sulahiri@stanford.edu}
%
}

%\nipsfinalcopy % Uncomment for camera-ready version

@@

@@

In this section we'll provide a summary of all the relevant properties of ergodic Markov chains in continuous time to define notation and .
It is a straightforward generalisation of material that can be found in \cite{kemeny1960finite} with some ideas from \cite{hunter2000survey}.

@@

This allows us to write the lumpability condition \eqref{eq:lump}, and the transition matrix for the lumped process compactly:
%
\begin{equation}\label{eq:lumpcompact}
  VU\MM V = \MM V,
  \qquad
  \widehat{\MM} = U\MM V.
\end{equation}
%
By induction, one can show that similar relations hold for all powers:
%
\begin{equation}\label{eq:lumppower}
  VU\MM^nV = \MM^nV,
  \qquad
  \widehat{\MM}^n = U\MM^n V,
\end{equation}
%
and, via the Taylor series, for the exponential as well:
%
\begin{equation}\label{eq:lumpexp}
  VU \e^{t\MM} V = \e^{t\MM}V,
  \qquad
  \e^{t\widehat{\MM}} = U\e^{t\MM} V.
\end{equation}
%
The equilibrium distribution of the lumped process is given by
%
\begin{equation}\label{eq:lumpeq}
  \widehat{\mathbf{p}}^\infty = \eq V.
\end{equation}
%
\section{Signal-to-Noise ratio (SNR)}\label{sec:SNR}

@@

The individual potentiation/depression events will be described by \emph{discrete}-time Markov chains:
%
\begin{equation}\label{eq:MWdef}
  \M\potdep \equiv \I + \W\potdep,
  \qquad
  \M\potdep\onev = \onev,
  \qquad
  \M\potdep_{ij} \in [0,1].
\end{equation}
%
The initial signal creation event occurs at time $t=0$, but all subsequent potentiation/depression events occur at random times according to Poisson processes with rates $rf\potdep$, where $f\pot +f\dep =1$ are the fraction of plasticity events that are potentiating/depressing respectively.
This means that the ``forgetting'' process will be described by the \emph{continuous}-time Markov chain:
%
\begin{equation}\label{eq:forgetting}
  \MM = r \frg \equiv r\prn{f\pot\W\pot + f\dep\W\dep}.
\end{equation}
%
We only require that this Markov chain is ergodic. The Markov chains described by $\M\potdep $ need not be.

We assume that the probability distribution starts in the equilibrium distribution \eqref{eq:equilibrium}.
During the initial signal creation, a fraction $f\pot$ will change to $\eq\M\pot$ and a fraction $f\dep$ will change to $\eq\M\dep$.
After this, probabilities will evolve according to \eqref{eq:rowsum}.

@@

As discussed in the main text, the signal-to-noise ratio is given by
%
\begin{equation}\label{eq:SNRcurve}
  \SNR(t) = \sqrt{N} (2f\pot f\dep ) \eq (\W\pot  - \W\dep )\e^{rt\frg}\w.
\end{equation}
%

We can express this in terms of the one parameter family of transition matrices:
%
\begin{equation}\label{eq:Walpha}
  \begin{aligned}
  \W(\alpha) &= \alpha\W\pot +(1-\alpha)\W\dep ,
  &\quad&\implies&
    \frg&=\W(f\pot ),\\&&&&
    \W\pot  - \W\dep  &= \diff{\W}{\alpha},\\&&&&
    \eq \diff{\W}{\alpha} &= -\diff{\eq}{\alpha} \frg.
  \end{aligned}
\end{equation}
%
Then \eqref{eq:SNRcurve} becomes
%
\begin{equation}\label{eq:SNRalpha}
  \SNR(t) = \sqrt{N} (2f\pot f\dep ) \diff{\eq}{\alpha}(-\frg)\,\e^{rt\frg}\w.
\end{equation}
%

@@

Suppose that we have a partition such that $\W\pot $ and $\W\dep $ are simultaneously lumpable, and that all the states in each subset have the same synaptic strength (see \sref{sec:lump}):
%
\begin{equation}\label{eq:lumpablesynapse}
  VU\W\potdep  V = \W\potdep  V,
  \qquad
  VU\w=\w.
\end{equation}
%
We can define a new synapse with
%
\begin{equation}\label{eq:lumpedsynapse}
  \widehat{\W}\potdep  = U\W\potdep  V,
  \qquad
  \widehat{\w} = U \w,
  \qquad
  \widehat{\mathbf{p}}^\infty = \eq V.
\end{equation}
%
This synapse has an SNR curve:
%
\begin{equation}\label{eq:lumpedSNR}
  \begin{aligned}
    \frac{\SNR(t)}{\sqrt{N} (2f\pot f\dep )} &=  \widehat{\mathbf{p}}^\infty (\widehat{\W}\pot  - \widehat{\W}\dep )\e^{rt\widehat{\W}^F}\widehat{\w}. \\
      &= \eq VU (\W\pot  - \W\dep ) VU \e^{rt\frg} VU \w. \\
      &= \eq (\W\pot  - \W\dep ) VU \e^{rt\frg} VU \w. \\
      &= \eq (\W\pot  - \W\dep )\e^{rt\frg} VU \w. \\
      &= \eq (\W\pot  - \W\dep )\e^{rt\frg}\w. \\
  \end{aligned}
\end{equation}
%
\ie the lumped process has exactly the same SNR as the original one.

@@

Using $\eq\frg=0$, we can write the initial SNR as
%
\begin{equation}\label{eq:init}
  \frac{\SNR(0)}{\sqrt{N}} = I = ({\eq}^++{\eq}^-)(f\pot \W\pot -f\dep \W\dep )(\onev^+-\onev^-).
\end{equation}
%
Using $\W\potdep (\onev^++\onev^-)=0$ and \eqref{eq:compflux}:
%
\begin{equation*}
  r{\eq}^-(f\pot \W\pot +f\dep \W\dep )\onev^+ = \F_{-+} = \F_{+-} = r{\eq}^+(f\pot \W\pot +f\dep \W\dep )\onev^-,
\end{equation*}
%
we can rewrite \eqref{eq:init} as
%
\begin{equation}\label{eq:initflux}
  I = \frac{4\F_{-+}}{r} - 4{\eq}^+f\pot \W\pot \onev^- - 4{\eq}^-f\dep \W\dep \onev^+.
\end{equation}
%
The last two terms are guaranteed to be negative, as the diagonal parts of $\W\potdep $ cannot contribute.
Therefore
%
\begin{equation}\label{eq:initfluxineq}
  \SNR(0) \leq \frac{4\sqrt{N}\F_{-+}}{r}.
\end{equation}
%
This inequality is saturated if potentiation never takes it from a $+$ state to a $-$ state and depression never takes it from a $-$ state to a $+$ state.

@@

The signal-to-noise curve is given by \eqref{eq:SNRalpha}.
%%
%\begin{equation}\label{eq:SNR}
%  \SNR(t) = \sqrt{N}(2f^+f\dep ) \diff{\eq}{\alpha} (-\frg) \e^{rt\frg} \w.
%\end{equation}
%%
The area is computed by integrating this
%
\begin{equation}\label{eq:area}
\begin{aligned}
  A &= \frac{\sqrt{N}(2f\pot f\dep )}{r} \diff{\eq}{\alpha} \brk{-\e^{rt\frg}}_0^\infty \w \\
    &= \frac{\sqrt{N}(2f\pot f\dep )}{r} \diff{\eq}{\alpha} (\I-\onev\eq) \w \\
    &= \frac{\sqrt{N}(2f\pot f\dep )}{r} \diff{\eq}{\alpha} \w.
\end{aligned}
\end{equation}
%

We can rewrite this using \eqref{eq:diffpT}, with $A=\sqrt{N}(2f\pot f\dep )\hat{A}$ and $\enc_{ij} \equiv \diff{\frg_{ij}}{\alpha}=\W\pot _{ij}-\W\dep _{ij}$
%
\begin{equation}\label{eq:areaT}
  \hat{A} = \sum_{i,j,k} \eq_i \enc_{ij} (\fptb_{ik} - \fptb_{jk}) \eq_k \w_k.
\end{equation}
%

@@

\subsection{Derivatives \wrt \texorpdfstring{$\W\potdep $}{W+/-}}\label{sec:deriv}

As discussed in the main text,
we will regard the off-diagonal elements of $\W\potdep _{ij}$ to be the independent variables,
with $\W\potdep _{ii}=-\sum_{j \neq i} \W\potdep _{ij}$ imposed by hand.
Thus,
%
\begin{equation}\label{eq:basicderivs}
  \pdiff{\frg_{ij}}{\W\potdep _{gh}} = f\potdep  \delta_{gi}(\delta_{hj}-\delta_{ij}),
  \qquad
  \pdiff{\enc_{ij}}{\W\potdep _{gh}} = \pm\delta_{gi}(\delta_{hj}-\delta_{ij}).
\end{equation}
%
The implicit $g \neq h$ that comes with all derivatives is unnecessary, as the derivatives above vanish when $g=h$.

In particular, differentiating \eqref{eq:funddef},
%
\begin{equation}\label{eq:derivZ}
  \pdiff{\fund_{ij}}{\W\potdep _{gh}} = rf\potdep  \fund_{ig} (\fund_{hj}-\fund_{gj}).
\end{equation}
%
We can then differentiate expression \eqref{eq:areaZ} to get
%
\begin{equation}\label{eq:derivA}
  \begin{aligned}
    \pdiff{\hat{A}}{\W\potdep _{gh}} =&\,
%      rf\potdep  \sum_{ijkl} \enc_{ij} \w_k \pib_l \brk{
%         \fund_{lg}(\fund_{hi}-\fund_{gi})(\fund_{jk}-\fund_{ik}) + \fund_{li}(\fund_{jg}-\fund_{ig})(\fund_{hk}-\fund_{gk}) } \\
%    &\pm \sum_{kl} \w_k \pib_l \fund_{lg} (\fund_{hk} - \fund_{gk}) \\
%    =&\,
%      2rf\potdep  \sum_{ij} \enc_{ij} \eq_i \eq_g \brk{
%         (\fptb_{gi}-\fptb_{hi})(\eta^+_{i}-\eta^+_{j}) +
%         (\fptb_{ig}-\fptb_{jg})(\eta^+_{g}-\eta^+_{h}) } \\
%      &\pm 2 \eq_g (\eta^+_g - \eta^+_h) \\
%    =&\,
%      2rf\potdep  \sum_{ij} \enc_{ij} \eq_i \eq_g
%         (\fptb_{gi}-\fptb_{hi})(\eta^+_{i}-\eta^+_{j})
%      \pm 2 \eq_g (\eta^+_g - \eta^+_h)\brk{1+\pdiff{\ln(\eq_g)}{\,\ln(f\potdep )}} \\
%    =&\,
      2rf\potdep  \eq_g \brk{\sum_i a_i (\fptb_{gi}-\fptb_{hi}) + c_g (\eta^+_{g}-\eta^+_{h})}
      \pm 2 \eq_g (\eta^+_g - \eta^+_h). \\
  \end{aligned}
\end{equation}
%
where $a_i$ and $c_k$ were defined in \eqref{eq:areacoeffs}.

It is sometimes useful to consider the following derivatives:
%
\begin{equation}\label{eq:derivqw}
  \begin{aligned}
    \pdiff{}{\frg_{gh}} &\equiv     \pdiff{}{\W\pot _{gh}} +     \pdiff{}{\W\dep _{gh}} ,&\qquad
    \pdiff{}{\enc_{gh}}    &\equiv f\dep  \pdiff{}{\W\pot _{gh}} - f\pot  \pdiff{}{\W\dep _{gh}} .
  \end{aligned}
\end{equation}
%
Each of these derivatives behaves as their names suggest:
%
\begin{equation}\label{eq:derivqweff}
  \pdiff{\frg_{ij}}{\frg_{gh}} = \pdiff{\enc_{ij}}{\enc_{gh}}
  =  \delta_{gi}(\delta_{hj}-\delta_{ij}),
  \qquad
  \pdiff{\enc_{ij}}{\frg_{gh}} = \pdiff{\frg_{ij}}{q_{gh}}  = 0.
\end{equation}
%
This is because we could treat $\frg$ and $q$ as the independent variables. However, the boundaries of the allowed region are more easily expressed in terms of $\W\potdep $.
